Safe Transferable Regions

### 2015-11-19 paper/related.tex
Edits.

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust, gpu14}. Our work differs from the existing proposals in a
number of ways. Firstly, our problem setting includes lexically scoped
stack regions and dynamic transferable regions (both
programmer-managed) in context of an object-oriented programming
language equipped with higher-order functions. Second, we adopt a
two-pronged approach to memory safety that relies on a combination of
a simple static type discipline and lightweight runtime checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach requires neither restrictive static
mechanisms (e.g., linear types and unique pointers) nor expensive
runtime mechanisms (e.g., garbage collection and reference counting)
in order to guarantee safety. Lastly, our region type system comes
equipped with full type inference that completely eliminates the need
to write region annotations on types to convince the type checker that
the program is safe.

~\cite{tofte97} proposes to extend Standard ML, a higher-order
functional language, with lexically-scoped stack regions, and defines
an elaboration from Standard ML to the region-annotated version of
Standard ML. The aim of the elaboration is to introduce stack regions
and do away with GC in a transparent fashion without jeopardizing
memory safety. We too define an elaboration, but our focus is on
introducing region annotations necessary to prove the safety of an
object-oriented program that already uses (stack and dynamic) regions.
Similar to their region inference algorithm, our region type inference
algorithm can deal with region-polymorphic recursion by computing
fixed points for recursive constraints. While their inference
algorithm only ever generates equality constraints, which can be
solved via unification, our type inference algorithm also generates
partial order outlives constraints, which are required to capture
subtle relationships between lifetimes of transferable regions and
stack regions. Consequently, our constraint solving algorithm is more
sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

@@

An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}. Like us, they too
assume a source language with programmer-managed memory regions, and
focus on proving safety of programs written in that language. Their
source language admits various kinds of regions in order to support
shared-memory concurrency. However, all their regions have
lexically-scoped lifetimes. In contrast, we admit regions with
dynamically determined lifetimes in order to support message-passing
concurrency. We borrow outlives relation from their formal
development, and our type system bears some similarities to theirs.
However, our language also admits parametric polymorphism (generics)
and higher-order functions, whose interaction is non-trivial in
context of a region type system. On the other hand, our type system
eschews the notions of region kinds and ownership, leading to a more
succinct formalization. Furthermore, we establish a type safety result
that formalizes its guarantees with respect to a well-defined
operational semantics. Like Cyclone, ~\cite{MIT03}'s language is
explicitly typed. Although there is some support for local type
inference, region types for methods and classes need to be written
explicitly. In contrast, we support full type inference that
eliminates any such need.

>>>>

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically enforce the safety of
their region-based memory management scheme in ML. Following their
seminal work, static type systems for safe region-based memory
management have been extensively studied in the context of various
languages and problem settings~\cite{cyclone02, cyclone04, yates99,
MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
existing proposals in a number of ways. Firstly, our problem setting
includes lexically scoped stack regions and dynamic transferable
regions (both programmer-managed) in context of an object-oriented
programming language equipped with higher-order functions. Second, we
adopt a two-pronged approach to memory safety that relies on a
combination of a simple static type discipline and lightweight runtime
checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach requires neither restrictive static
mechanisms (e.g., linear types and unique pointers) nor expensive
runtime mechanisms (e.g., garbage collection and reference counting)
in order to guarantee safety. Lastly, our region type system comes
equipped with full type inference that more or less eliminates the
need to write region annotations on types to convince the type checker
that the program is safe.

~\cite{tofte97} proposes to extend Standard ML, a higher-order
functional language, with lexically-scoped stack regions, and defines
an elaboration from Standard ML to the region-annotated version of
Standard ML. The aim of the elaboration is to introduce stack regions
and do away with GC in a transparent fashion without jeopardizing
memory safety. We too define an elaboration, but our focus is on
introducing region annotations necessary to prove the safety of an
object-oriented program that already uses (stack and dynamic) regions.
% Similar to their region inference algorithm, our region type inference
% algorithm can deal with region-polymorphic recursion by computing
% fixed points for recursive constraints. 
Our type inference is analogous to their region inference. While their
inference algorithm only ever generates equality constraints, which
can be solved via unification, our type inference algorithm also
generates partial order outlives constraints, which are required to
capture subtle relationships between lifetimes of transferable regions
and stack regions. Consequently, our constraint solving algorithm is
more sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

@@

An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}.  
% Like us, they too assume a source language with programmer-managed
% memory regions, and focus on proving safety of programs written in
% that language.  
Their source language admits various kinds of regions with
lexically-scoped lifetimes (i.e., stack regions) in order to support
shared-memory concurrency. In contrast, we admit regions with
dynamically determined lifetimes in order to support message-passing
concurrency. We borrow outlives relation from their formal
development, and our type system bears some similarities to theirs.
However, our language also admits parametric polymorphism (generics)
and higher-order functions.
%, whose interaction is non-trivial in context of a region type system. 
% On the other hand, our type system eschews the notions of region kinds
% and ownership, leading to a more succinct formalization. 
Furthermore, we establish type safety and transfer safety results that
formalize the guarantees provided by our system. While ~\cite{MIT03}'s
language is explicitly typed, our language comes equipped with full
type inference.
% Although there is some support for local
% type inference, region types for methods and classes need to be
% written explicitly. In contrast, our region type inference that
% eliminates any such need.
